Short-time simulation of collective spin Hamiltonians with arbitrary decoherence

### 2018-11-04 soc_squeezing/cumulant_expansions.tex
changed convention for decoherence vector ordering to PZM changed convention (involving factors of 2) for dephasing rate for consistency with literature
added todo line into the cumulant expansions text
change section on the general commutator of spin operators to a section on the general product

Previous work has shown how to realize spin-squeezing Hamiltonians
with a variety of ultracold atomic and trapped ion systems.  While
simulating the dynamics of initially Gaussian states evolving under
these Hamiltonians is straightforward in the context of a closed
quantum system, here we develop a method to simulate short-time
dynamics in the presence of decoherence.  In particular, we will
consider dynamics induced by the one-axis twisting (OAT), two-axis
twisting (TAT), and transverse field (TVF) Hamiltonians of the form
\begin{align}
  H_{\t{OAT}} = \chi S_\z^2,
  &&
  H_{\t{TAT}}^{\z,\y} = \chi \p{S_\z^2 - S_\y^2},
  &&
  H_{\t{TVF}} = \chi S_\z^2 + \Omega S_\x.
\end{align}
In order to determine the squeezing behavior of these Hamiltonians, we
will study the dynamics of Heisenburg operators $\O$, whose evolution
under a Hamiltonian $H$ is given by
\begin{align}
  \f{d}{dt} \O = i\sp{H,\O}_- + \sum_\J \gamma_\J \D\p{\J;\O},
  &&
  \D\p{\J;\O}
  \equiv \sum_{J\in\J}\p{J^\dag \O J - \f12\sp{J^\dag J,\O}_+},
  \label{eq:EOM}
\end{align}
where $\J$ is a set of jump operators with a corresponding decoherence
rate $\gamma_\J$ and $\sp{X,Y}_\pm\equiv XY\pm YX$.  Our measure of
spin squeezing is
\begin{align}
  \xi^2 \equiv \f{N}{\abs{\bk{\v S}}^2}
  \min_{\uv n\perp\bk{\v S}} \bk{\p{\v S\c\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $N$ is the total number of spins, $\v S$ is the collective spin
vector, and the minimization is performed over all unit vectors
$\uv n$ in the plane orthogonal to the mean spin vector $\bk{\v S}$.
This squeezing parameter is entirely determined by components
$\bk{S_\alpha}$ of the mean spin vector in addition to collective
spin-spin correlators of the form $\bk{S_\alpha S_\beta}$; our task is
therefore to compute these expectation values in all scenarios of
interest.

@@

Ultimately, we are interested in the dynamics of collective spin
operators.  Choosing a basis $\set{\S_{\v m}}$ for these operators,
e.g.~$\S_{\v m}\equiv S_+^{m_+} S_\z^{m_\z} S_-^{m_-}$ with
$\v m\equiv\p{m_+,m_\z,m_-}\in\mathbb{N}_0^3$, we can expand all
collective-spin Hamiltonians in the form
\begin{align}
  H = \sum_{\v m} h_{\v m} \S_{\v m},
  \label{eq:H_general}
\end{align}
where if $\S_{\v m}=S_+^{m_+} S_\z^{m_\z} S_-^{m_-}$ then Hermiticity
requires that $h_{m_+,m_\z,m_-}=h_{m_-,m_\z,m_+}^*$.  The one-axis
twisting Hamiltonian with a transverse field, for example, is
\begin{align}
  H_{\t{TVF}}
  = \chi S_\z^2 + \Omega S_\x
  = \chi S_\z^2 + \f{\Omega}{2} S_+ + \f{\Omega}{2} S_-,
\end{align}
while the two-axis twisting Hamiltonians about $\p{z,y}$ and $\p{y,x}$
are
\begin{align}
  H_{\t{TAT}}^{\z,\y}
  \propto S_\z^2 - S_\y^2
  = S_\z^2 + \f14 S_+^2 + \f14 S_-^2
  - \f12 S_+ S_- + \f12 S_\z,
\end{align}
and
\begin{align}
  H_{\t{TAT}}^{\y,\x}
  \propto S_\y^2 - S_\x^2
  = -\f12 S_+^2 - \f12 S_-^2.
\end{align}
The evolution of a general operator $\S_{\v m}$ under a Hamiltonian of
the form in \eqref{eq:H_general} is given by
\begin{align}
  \f{d}{dt} \S_{\v m}
  = i \sum_{\v n} h_{\v n}
  \sp{\S_{\v n}, \S_{\v m}}_-
  + \sum_\J \gamma_\J \D\p{\J; \S_{\v m}}
  \equiv \sum_{\v n} T_{\v m\v n} \S_{\v n},
  \label{eq:time_derivative}
\end{align}
where commutator and decoherence terms necessary for computing the
matrix elements $T_{\v m\v n}$ of the time derivative operator $d/dt$
are worked out Appendices \ref{sec:comm_general},
\ref{sec:decoherence_single}, and \ref{sec:decoherence_collective}.

@@

\section{Basic spin operator identities}
\label{sec:identities}

@@

\section{Commutation relations between ordered products of collective
  spin operators}
\label{sec:comm_general}

We will generally be concerned with ordered products of collective
spin operators of the form $S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n$ for
$\bmu\equiv-\mu\in\set{+1,-1}$.  The most general (anti-)commutator we
need to consider is then
\begin{align}
  \sp{S_\mu^p \p{\mu S_\z}^q S_\bmu^r,
    S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n}_\pm
  = \S^{pqr}_{\ell mn;\mu} \pm \S^{\ell mn}_{pqr;\mu},
\end{align}
where
\begin{align}
  \S^{pqr}_{\ell mn;\mu}
  = S_\mu^p \p{\mu S_\z}^q S_\bmu^r
  S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n
  = \sum_{k=0}^{\min\set{r,\ell}} k! { r \choose k } { \ell \choose k }
  S_\mu^{p+\ell-k} \tilde Z_{qr\ell m;\mu}^{(k)} S_\bmu^{r+n-k},
  \label{eq:general_product}
\end{align}
for
\begin{align}
  \tilde Z_{qr\ell m;\mu}^{(k)}
  &\equiv \p{\ell-k+\mu S_\z}^q
  Z_{r\ell;\mu}^{(k)} \p{r-k+\mu S_\z}^m \\
  &= \sum_{a=0}^k \zeta_{r\ell}^{ka}
  \sum_{b=0}^q \p{\ell-k}^{q-b} { q \choose b }
  \sum_{c=0}^m \p{r-k}^{m-c} { m \choose c }
  \p{\mu S_\z}^{a+b+c},
\end{align}
defined in terms of $Z_{r\ell;\mu}^{(k)}$ and $\zeta_{r\ell}^{ka}$ as
given in \eqref{eq:Z_mnp} and \eqref{eq:zeta_mnpq}.

@@

The effect of decoherence via single-spin dephasing is described by
\begin{align}
  \D\p{\sigma_\z/\sqrt2; \O}
  = \f12 \sum_j\p{\sigma_\z^{(j)} \O \sigma_\z^{(j)}
    - \f12\sp{\sigma_\z^{(j)} \sigma_\z^{(j)},\O}_+}
  = \f12 \sum_j \sigma_\z^{(j)} \O \sigma_\z^{(j)} - S \O,
\end{align}
where the factor $\sqrt2$ is inserted as
e.g.~$\D\p{\sigma_\z/\sqrt2;S_\mu}=-S_\mu$, such that the above
expression generates decay of single-weight transverse collective spin
operators at a unit rate.  From \eqref{eq:san_z_z}, we then have
\begin{align}
  \D\p{\sigma_\z/\sqrt2; S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n}
  = -\p{\ell+n} S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n
  + 2 \ell n S_\mu^{\ell-1} \p{S + \mu S_\z}
  \p{-1 + \mu S_\z}^m S_\bmu^{n-1}.
\end{align}
Decoherence via single-spin dephasing makes operators
$S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n$ ``grow'' in $m$, although the sum
$\ell+m+n$ does not grow.

@@

  S_\mu^{\ell+1} \p{\mu S_\z}

@@

  = \sum_{\tilde{\v\rho}} g_{\tilde{\v\rho}} \v\eta_{\tilde{\v\rho}}

>>>>

Previous work has shown how to realize spin-squeezing Hamiltonians
with a variety of ultracold atomic and trapped ion systems.  While
simulating the dynamics of initially Gaussian states evolving under
these Hamiltonians is straightforward in the context of a closed
quantum system, here we develop a method to simulate short-time
dynamics in the presence of decoherence.  In particular, we will
consider dynamics induced by the one-axis twisting (OAT), two-axis
twisting (TAT), and transverse field (TVF) Hamiltonians of the form
\begin{align}
  H_{\t{OAT}} = \chi S_\z^2,
  &&
  H_{\t{TAT}}^{\z,\y} = \chi \p{S_\z^2 - S_\y^2},
  &&
  H_{\t{TVF}} = \chi S_\z^2 + \Omega S_\x.
\end{align}
In order to determine the squeezing behavior of these Hamiltonians, we
will study the dynamics of Heisenburg operators $\O$, whose evolution
under a Hamiltonian $H$ is given by
\begin{align}
  \f{d}{dt} \O = i\sp{H,\O}_- + \sum_\J \gamma_\J \D\p{\J;\O},
  &&
  \D\p{\J;\O}
  \equiv \sum_{J\in\J}\p{J^\dag \O J - \f12\sp{J^\dag J,\O}_+},
  \label{eq:EOM}
\end{align}
where $\J$ is a set of jump operators with a corresponding decoherence
rate $\gamma_\J$ and $\sp{X,Y}_\pm\equiv XY\pm YX$.  Our measure of
spin squeezing is
\begin{align}
  \xi^2 \equiv \f{N}{\abs{\bk{\v S}}^2}
  \min_{\uv n\perp\bk{\v S}} \bk{\p{\v S\c\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $N$ is the total number of spins, $\v S$ is the collective spin
vector, and the minimization is performed over all unit vectors
$\uv n$ in the plane orthogonal to the mean spin vector $\bk{\v S}$.
This squeezing parameter is entirely determined by one- and two-spin
correlators of the form $\bk{S_\alpha}$ and $\bk{S_\alpha S_\beta}$;
our task is therefore to compute these expectation values in all
scenarios of interest.

@@

Ultimately, we are interested in the dynamics of collective spin
operators.  Choosing a basis $\set{\S_{\v m}}$ for these operators,
e.g.~$\S_{\v m}\equiv S_+^{m_+} S_\z^{m_\z} S_-^{m_-}$ with
$\v m\equiv\p{m_+,m_\z,m_-}\in\mathbb{N}_0^3$, we can expand all
collective-spin Hamiltonians in the form
\begin{align}
  H = \sum_{\v m} h_{\v m} \S_{\v m},
  \label{eq:H_general}
\end{align}
where if $\S_{\v m}=S_+^{m_+} S_\z^{m_\z} S_-^{m_-}$ then Hermiticity
requires that $h_{m_+,m_\z,m_-}=h_{m_-,m_\z,m_+}^*$.  The one-axis
twisting Hamiltonian with a transverse field, for example, is
\begin{align}
  H_{\t{TVF}}
  = \chi S_\z^2 + \Omega S_\x
  = \chi S_\z^2 + \f{\Omega}{2} S_+ + \f{\Omega}{2} S_-,
\end{align}
while the two-axis twisting Hamiltonians about $\p{z,y}$ and $\p{y,x}$
are
\begin{align}
  H_{\t{TAT}}^{\z,\y}
  \propto S_\z^2 - S_\y^2
  = S_\z^2 + \f14 S_+^2 + \f14 S_-^2
  - \f12 S_+ S_- + \f12 S_\z,
\end{align}
and
\begin{align}
  H_{\t{TAT}}^{\y,\x}
  \propto S_\y^2 - S_\x^2
  = -\f12 S_+^2 - \f12 S_-^2.
\end{align}
The evolution of a general operator $\S_{\v m}$ under a Hamiltonian of
the form in \eqref{eq:H_general} is given by
\begin{align}
  \f{d}{dt} \S_{\v m}
  = i \sum_{\v n} h_{\v n}
  \sp{\S_{\v n}, \S_{\v m}}_-
  + \sum_\J \gamma_\J \D\p{\J; \S_{\v m}}
  \equiv \sum_{\v n} T_{\v m\v n} \S_{\v n},
  \label{eq:time_derivative}
\end{align}
where commutator and decoherence terms necessary for computing the
matrix elements $T_{\v m\v n}$ of the time derivative operator $d/dt$
are worked out Appendices \ref{sec:prod_general},
\ref{sec:decoherence_single}, and \ref{sec:decoherence_collective}.

@@

\section{Basic spin operator identities}
\label{sec:identities}
% TODO: change { \sigma_\mu } --> { s_\mu }

@@

\section{Product of arbitrary ordered collective spin operators}
\label{sec:prod_general}

The most general product of collective spin operators that we need to
compute is
\begin{align}
  \S^{pqr}_{\ell mn;\mu}
  = S_\mu^p \p{\mu S_\z}^q S_\bmu^r
  S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n
  = \sum_{k=0}^{\min\set{r,\ell}} k! { r \choose k } { \ell \choose k }
  S_\mu^{p+\ell-k} \tilde Z_{qr\ell m;\mu}^{(k)} S_\bmu^{r+n-k},
  \label{eq:general_product}
\end{align}
where
\begin{align}
  \tilde Z_{qr\ell m;\mu}^{(k)}
  &\equiv \p{\ell-k+\mu S_\z}^q
  Z_{r\ell;\mu}^{(k)} \p{r-k+\mu S_\z}^m \\
  &= \sum_{a=0}^k \zeta_{r\ell}^{ka}
  \sum_{b=0}^q \p{\ell-k}^{q-b} { q \choose b }
  \sum_{c=0}^m \p{r-k}^{m-c} { m \choose c }
  \p{\mu S_\z}^{a+b+c},
\end{align}
is defined in terms of $Z_{r\ell;\mu}^{(k)}$ and $\zeta_{r\ell}^{ka}$
as respectively given in \eqref{eq:Z_mnp} and \eqref{eq:zeta_mnpq}.
The (anti-)commutator of two ordered products of collective spin
operators is then
\begin{align}
  \sp{S_\mu^p \p{\mu S_\z}^q S_\bmu^r,
    S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n}_\pm
  = \S^{pqr}_{\ell mn;\mu} \pm \S^{\ell mn}_{pqr;\mu}.
\end{align}

@@

The effect of decoherence via single-spin dephasing is described by
\begin{align}
  \D\p{\sigma_\z; \O}
  = \sum_j\p{\sigma_\z^{(j)} \O \sigma_\z^{(j)}
    - \f12\sp{\sigma_\z^{(j)} \sigma_\z^{(j)},\O}_+}
  = \sum_j \sigma_\z^{(j)} \O \sigma_\z^{(j)} - 2 S \O.
\end{align}
From \eqref{eq:san_z_z}, we then have
\begin{align}
  \D\p{\sigma_\z; S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n}
  = -2\p{\ell+n} S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n
  + 4 \ell n S_\mu^{\ell-1} \p{S + \mu S_\z}
  \p{-1 + \mu S_\z}^m S_\bmu^{n-1}.
\end{align}
Decoherence via single-spin dephasing makes operators
$S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n$ ``grow'' in $m$, although the sum
$\ell+m+n$ does not grow.

@@

  S_\mu^{\ell+1} \mu S_\z

@@
